---
title: "BART-CMA Applied Mediation: Mixture → DII → Liver Markers"
author: "Augustina"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
params:
  profile: "test"                       # "test" (RStudio) or "production" (SLURM batch render)
  data_xlsx: "completed_liver_pfas_ALL.xlsx"
  confounder_set: "primary"             # "primary" (recommended) or "sensitivity" (over-adjusted)
  log_outcomes: true
  weight_col: "PFASWeight"              # NULL/"" for an unweighted sample-level analysis
  n_cores: 1                            # >1 only in a batch job, never in interactive RStudio
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = FALSE, fig.width = 9, fig.height = 6)

## Allows running chunks interactively (no knit) without a params object.
if (!exists("params")) {
  params <- list(profile = "test",
                 data_xlsx = "completed_liver_pfas_ALL.xlsx",
                 confounder_set = "primary", log_outcomes = TRUE,
                 weight_col = "PFASWeight", n_cores = 1)
}
```

# 1. Packages

```{r packages}
suppressPackageStartupMessages({
  library(BART)        # wbart / pbart
  library(readxl)      # read the 20 imputation sheets
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(writexl)
  library(parallel)    # mclapply for the batch (production) run
})
```

# 2. Configuration

Everything that defines the analysis lives here. The **only scientific choice
you must own is the confounder set** (see the note below).

```{r config}
CFG <- list(
  exposure_cols = c("log_Lead", "log_Cadmium", "log_Mercury",
                    "log_PFOA", "log_PFOS"),
  mediator_col  = "DII",
  outcome_cols  = c("LiverFibrosis", "LiverFat", "ALT", "AST",
                    "AlkPhosphatase", "GGT"),
  outcome_type  = "continuous",          # all six are continuous here
  weight_col    = if (is.null(params$weight_col) || params$weight_col == "")
                    NULL else params$weight_col,
  log_outcomes  = isTRUE(params$log_outcomes),

  ## PRIMARY = pre-exposure common causes only. Drops CRP, HbA1c, lipids, BP,
  ## waist — these sit DOWNSTREAM of DII/exposures (mediators/colliders), and
  ## conditioning on them biases NIE toward 0.
  confounders_primary = c("Age", "BMI", "Gender_Male", "Race_NHWhite",
                          "Education", "Income", "Smoker", "Drinker"),
  ## SENSITIVITY = your original over-adjusted set; report alongside PRIMARY.
  confounders_sensitivity = c("Age", "BMI", "Gender_Male", "Race_NHWhite",
                              "Education", "Income", "Smoker", "Drinker",
                              "WaistCircumference", "CRP", "HbA1c",
                              "TotalCholesterol", "HDLCholesterol",
                              "SystolicBloodPressure", "DiastolicBloodPressure")
)
CFG$confounder_cols <- if (params$confounder_set == "sensitivity")
  CFG$confounders_sensitivity else CFG$confounders_primary

## ---- profile: test (RStudio dev) vs production (SLURM batch) ---------------
if (params$profile == "production") {
  RUN <- list(outcomes = CFG$outcome_cols, n_imp = 20,
              n_tree = 200, n_burn = 2000, n_post = 2000, n_med = 30)
} else {
  RUN <- list(outcomes = c("ALT"), n_imp = 3,        # light, finishes in minutes
              n_tree = 50, n_burn = 500, n_post = 500, n_med = 10)
}

cat("Profile        :", params$profile, "\n")
cat("Outcomes       :", paste(RUN$outcomes, collapse = ", "), "\n")
cat("Imputations    :", RUN$n_imp, "\n")
cat("Confounder set :", params$confounder_set,
    "(", length(CFG$confounder_cols), "vars )\n")
cat("Weighted       :", !is.null(CFG$weight_col),
    if (!is.null(CFG$weight_col)) paste0("(", CFG$weight_col, ")") else "", "\n")
cat("Log outcomes   :", CFG$log_outcomes, "\n")
cat("Total fits     :", length(RUN$outcomes) * RUN$n_imp, "\n")
```

> **Confounder decision.** Defaulting to `primary`. CRP is downstream systemic
> inflammation on the DII→liver path; conditioning on it deflates the very NIE
> you're estimating. Run `sensitivity` too and report both — if NIE shrinks
> under the over-adjusted set, that's evidence of pathway over-adjustment.

# 3. Load imputed datasets (20 sheets → list)

```{r load-data}
sheets <- excel_sheets(params$data_xlsx)
sheets <- sheets[grepl("^Imputation_", sheets)]
completed_list <- lapply(sheets, function(sh)
  as.data.frame(read_excel(params$data_xlsx, sheet = sh)))
names(completed_list) <- sheets

cat("Loaded", length(completed_list), "imputed datasets:",
    nrow(completed_list[[1]]), "rows x", ncol(completed_list[[1]]), "cols.\n")

## column check
need <- c(CFG$exposure_cols, CFG$mediator_col, RUN$outcomes,
          CFG$confounder_cols, CFG$weight_col)
miss <- setdiff(need, colnames(completed_list[[1]]))
if (length(miss)) { cat("MISSING columns:\n"); print(miss) } else
  cat("All required columns present.\n")
```

# 4. BART-CMA estimator (Devick 2022 framework)

Corrected vs. the earlier n200 version: `predict()` is out of the posterior
loop (≈50–200× faster), `sigma` is read from the post-burn-in draws, survey
weights enter via a weighted population average, and there's a `pbart` branch
for binary outcomes (unused here, kept for reuse).

```{r bart-cma-function}
bart_cma_mixture <- function(Z, M, Y, X, weights = NULL,
                             q_low = 0.25, q_high = 0.75,
                             outcome_type = c("continuous", "binary"),
                             n_tree = 200, n_burn = 2000, n_post = 2000,
                             n_med_draws = 30, seed = 2025) {
  outcome_type <- match.arg(outcome_type)
  set.seed(seed)
  n <- length(Y); S <- n_post

  if (is.null(weights)) weights <- rep(1, n)
  weights <- as.numeric(weights); wsum <- sum(weights)
  wavg <- function(mat) as.numeric(mat %*% weights) / wsum

  Z <- as.matrix(Z); X <- as.matrix(X)
  Z_low  <- apply(Z, 2, quantile, probs = q_low,  na.rm = TRUE)
  Z_high <- apply(Z, 2, quantile, probs = q_high, na.rm = TRUE)
  Z_at_low  <- matrix(rep(Z_low,  each = n), nrow = n); colnames(Z_at_low)  <- colnames(Z)
  Z_at_high <- matrix(rep(Z_high, each = n), nrow = n); colnames(Z_at_high) <- colnames(Z)

  ## Step 1: mediator model E[M | Z, X]
  mfit <- wbart(x.train = cbind(Z, X), y.train = M,
                ntree = n_tree, nskip = n_burn, ndpost = n_post,
                printevery = n_burn + n_post + 1L)
  M_low_hat  <- predict(mfit, newdata = cbind(Z_at_low,  X))
  M_high_hat <- predict(mfit, newdata = cbind(Z_at_high, X))
  L <- length(mfit$sigma)
  sig <- mfit$sigma[(L - n_post + 1):L]          # POST-burn-in residual SDs

  ## Step 2: outcome model E[Y | Z, M, X]
  if (outcome_type == "continuous") {
    ofit <- wbart(x.train = cbind(Z, M, X), y.train = Y,
                  ntree = n_tree, nskip = n_burn, ndpost = n_post,
                  printevery = n_burn + n_post + 1L)
    predict_Y <- function(nd) predict(ofit, newdata = nd)
  } else {
    ofit <- pbart(x.train = cbind(Z, M, X), y.train = Y,
                  ntree = n_tree, nskip = n_burn, ndpost = n_post,
                  printevery = n_burn + n_post + 1L)
    predict_Y <- function(nd) predict(ofit, newdata = nd)$prob.test
  }

  ## Step 3: paired-draw Monte-Carlo g-computation
  ## Mediator draw s is paired with outcome draw s; residual realisations
  ## are averaged within the draw. The previous version drew one mediator
  ## index per k and shared it across all S rows, which averaged mediator
  ## uncertainty out of the posterior and produced intervals that were too
  ## narrow. Survey weighting (wavg) is unchanged.
  S_use   <- min(S, 200L)
  idx     <- if (S_use < S) round(seq(1, S, length.out = S_use)) else seq_len(S)
  n_med_paired <- 5L

  Y_hh <- Y_ll <- Y_hl <- matrix(NA_real_, length(idx), n)
  ridx  <- rep(seq_len(n), n_med_paired)
  Zh_r  <- Z_at_high[ridx, , drop = FALSE]
  Zl_r  <- Z_at_low[ridx,  , drop = FALSE]
  X_r   <- X[ridx, , drop = FALSE]
  B     <- n_med_paired * n
  cm    <- function(v) colMeans(matrix(v, n_med_paired, n, byrow = TRUE))

  for (j in seq_along(idx)) {
    sdx    <- idx[j]
    m_low  <- as.vector(t(matrix(M_low_hat[sdx, ],  n_med_paired, n, byrow = TRUE) +
                          matrix(rnorm(B, 0, sig[sdx]), n_med_paired, n)))
    m_high <- as.vector(t(matrix(M_high_hat[sdx, ], n_med_paired, n, byrow = TRUE) +
                          matrix(rnorm(B, 0, sig[sdx]), n_med_paired, n)))
    Y_hh[j, ] <- cm(predict_Y(cbind(Zh_r, M = m_high, X_r))[sdx, ])
    Y_ll[j, ] <- cm(predict_Y(cbind(Zl_r, M = m_low,  X_r))[sdx, ])
    Y_hl[j, ] <- cm(predict_Y(cbind(Zh_r, M = m_low,  X_r))[sdx, ])
  }

  TE  <- wavg(Y_hh) - wavg(Y_ll)
  NDE <- wavg(Y_hl) - wavg(Y_ll)
  NIE <- wavg(Y_hh) - wavg(Y_hl)
  PM  <- ifelse(abs(TE) > 1e-8, NIE / TE, NA_real_)
  list(TE_draws = TE, NDE_draws = NDE, NIE_draws = NIE, PM_draws = PM)
}
```

# 5. Run all (outcome × imputation) cells

```{r run, results='hold'}
## build the task grid
grid <- expand.grid(outcome = RUN$outcomes, imp = seq_len(RUN$n_imp),
                    stringsAsFactors = FALSE)

run_cell <- function(i) {
  outcome <- grid$outcome[i]; imp <- grid$imp[i]
  d <- completed_list[[imp]]
  num <- function(df) {
    df <- as.data.frame(df)
    df[] <- lapply(df, function(col)
      if (is.numeric(col)) col else as.numeric(as.character(col)))
    as.matrix(df)
  }
  Z <- num(d[, CFG$exposure_cols, drop = FALSE])
  M <- as.numeric(as.character(d[[CFG$mediator_col]]))
  Y <- as.numeric(as.character(d[[outcome]]))
  if (CFG$log_outcomes) { stopifnot(all(Y > 0)); Y <- log(Y) }
  conf <- intersect(CFG$confounder_cols, colnames(d))
  X <- num(d[, conf, drop = FALSE])
  w <- if (!is.null(CFG$weight_col)) d[[CFG$weight_col]] else NULL
  res <- bart_cma_mixture(Z, M, Y, X, weights = w,
                          outcome_type = CFG$outcome_type,
                          n_tree = RUN$n_tree, n_burn = RUN$n_burn,
                          n_post = RUN$n_post, n_med_draws = RUN$n_med,
                          seed = 2025 + i)
  list(outcome = outcome, imp = imp, result = res)
}

t0 <- Sys.time()
if (params$n_cores > 1) {
  cells <- mclapply(seq_len(nrow(grid)), run_cell, mc.cores = params$n_cores)
} else {
  cells <- lapply(seq_len(nrow(grid)), function(i) {
    cat(sprintf("[%d/%d] %s imp %d\n", i, nrow(grid),
                grid$outcome[i], grid$imp[i])); run_cell(i)
  })
}
cat("Total runtime:",
    round(as.numeric(difftime(Sys.time(), t0, units = "mins")), 1), "min\n")
```

# 6. Pool posteriors across imputations

Mixture-of-posteriors pooling: concatenate the per-imputation posterior draws
of each estimand, then take quantiles — this propagates both imputation and
posterior uncertainty.

```{r pool}
summ <- function(d, outcome, name) {
  d <- d[is.finite(d)]
  data.frame(outcome = outcome, estimand = name,
             mean = mean(d), sd = sd(d),
             lower = unname(quantile(d, 0.025)),
             upper = unname(quantile(d, 0.975)),
             prob_pos = mean(d > 0), row.names = NULL)
}

pooled <- lapply(unique(grid$outcome), function(oc) {
  these <- Filter(function(x) x$outcome == oc, cells)
  grab <- function(nm) unlist(lapply(these, function(x) x$result[[nm]]))
  bind_rows(summ(grab("TE_draws"),  oc, "TE"),
            summ(grab("NDE_draws"), oc, "NDE"),
            summ(grab("NIE_draws"), oc, "NIE"),
            summ(grab("PM_draws"),  oc, "PM"))
})
results_table <- bind_rows(pooled)
```

# 7. Results table

```{r results-table}
scale_note <- if (CFG$log_outcomes) " (log scale)" else ""
cat("Mixture: Lead, Cadmium, Mercury, PFOA, PFOS (25th→75th pctl)",
    scale_note, "\nMediator: DII | Confounders:", params$confounder_set,
    "| Weighted:", !is.null(CFG$weight_col), "\n\n")

results_table %>%
  mutate(across(c(mean, sd, lower, upper), ~round(., 4)),
         prob_pos = round(prob_pos, 3),
         CI = paste0("[", lower, ", ", upper, "]")) %>%
  select(outcome, estimand, mean, sd, CI, prob_pos) %>%
  as.data.frame() %>% print.data.frame(row.names = FALSE)
```

# 8. Forest plot — NIE

```{r forest-nie, fig.height=5}
nie <- results_table %>% filter(estimand == "NIE")
ggplot(nie, aes(mean, reorder(outcome, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = .2, linewidth = .8) +
  geom_point(size = 3, color = "darkblue") +
  labs(title = "Natural Indirect Effect (NIE) through DII",
       subtitle = paste0("Mixture 25th→75th pctl shift", scale_note),
       x = "NIE", y = NULL) +
  theme_minimal(base_size = 13) + theme(plot.title = element_text(face = "bold"))
```

# 9. Forest plot — TE

```{r forest-te, fig.height=5}
te <- results_table %>% filter(estimand == "TE")
ggplot(te, aes(mean, reorder(outcome, mean))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = .2, linewidth = .8) +
  geom_point(size = 3, color = "darkred") +
  labs(title = "Total Effect (TE) of mixture on liver markers",
       subtitle = paste0("Mixture 25th→75th pctl shift", scale_note),
       x = "TE", y = NULL) +
  theme_minimal(base_size = 13) + theme(plot.title = element_text(face = "bold"))
```

# 10. Proportion mediated

```{r pm, fig.height=5}
pm <- results_table %>% filter(estimand == "PM") %>% mutate(pct = mean * 100)
ggplot(pm, aes(reorder(outcome, pct), pct)) +
  geom_col(fill = "steelblue", alpha = .8) +
  geom_text(aes(label = paste0(round(pct, 1), "%")), hjust = -0.2, size = 4) +
  coord_flip() +
  labs(title = "Proportion mediated through DII", x = NULL, y = "% mediated") +
  theme_minimal(base_size = 13) + theme(plot.title = element_text(face = "bold")) +
  ylim(min(pm$pct) - 5, max(pm$pct) + 12)
```

# 11. Save outputs

```{r save}
dir.create("output", showWarnings = FALSE)
tag <- paste0(params$profile, "_", params$confounder_set)
saveRDS(cells,          file.path("output", paste0("bart_cma_cells_", tag, ".rds")))
saveRDS(results_table,  file.path("output", paste0("bart_cma_pooled_", tag, ".rds")))
write_xlsx(results_table, file.path("output", paste0("bart_cma_pooled_", tag, ".xlsx")))
cat("Saved to output/ with tag:", tag, "\n")
```

---

### How this Rmd is meant to be run

- **In RStudio Server (development):** keep `profile: test`. Knit or run chunks;
  finishes in a few minutes on 1 outcome × 3 imputations at reduced settings.
- **Full run (production):** do **not** knit this interactively. Render it as a
  batch job (see the accompanying instructions) with
  `params = list(profile = "production", n_cores = 10)`.
